 

### 2023-06-26 markdown/Thesis.tex
cite file added

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
2cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{sectsty}
\usepackage{paralist}
\usepackage{setspace}\spacing{1.5}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{dcolumn}
\usepackage{natbib}\bibliographystyle{agsm}
\usepackage[nottoc, numbib]{tocbibind}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

(the spacing is set to 1.5)

no more than 250 words for the abstract

\begin{itemize}
\tightlist
\item
  a description of the research question/knowledge gap -- what we know
  and what we don't know
\item
  how your research has attempted to fill this gap
\item
  a brief description of the methods
\item
  brief results
\item
  key conclusions that put the research into a larger context
\end{itemize}

@@

Thank you for following this tutorial!

I hope you'll find it useful to write a very professional dissertation.

@@

@@

The proteome is referred to the sum of all proteins of a given sample at
a given time. In the past several quantitative and qualitative assays
were used to enlighten the protein composition of a sample.

An early approach of qualitative analysis of the cellular proteome
involved labeling with fluorescent antibodies and imaging. The major
disadvantage of this technique was the limitation to only stain a few
proteins per cell. For quantification procedures such as single-cell
Western blots, immunoassays or CyTOF have been used. Other disadvantages
are the ability to permeate cells, accessibility and binding of the
epitope and the creation of specific antibodies for a given protein
\citep{Budnik2018}.

One of those techniques involved RNA-sequencing. Since RNA involves also
non-coding RNA, the amount of RNA is often not proportional to the
content of proteins in a cell. So the proteinaceous content of a cell
was only predicted and quantitative analysis was not possible.

\hypertarget{mass-spectrometry}{%
\subsection{Mass Spectrometry}\label{mass-spectrometry}}

Mass spectrometry enables qualitative and quantitative analysis of the
entire repertoire of a biological sample. The availability of gene
sequences in databases and the ability to match proteins against those
sequences with computational methods makes it possible to identify
alterations of a sample on a protein level. These alterations can rely
on the sequence level or could be to post-translational modifications
(PTMs) such as phosphorylation, methylation or else
\citep{Aebersold2003}.

Mass to charge ratio (m/z)

\pagenumbering{arabic}

@@

\hypertarget{materials}{%
\subsection{Materials}\label{materials}}

For analysis two types of cells were used. One type is the Jurkat-based
cell line (J-lat) with integrated HIV.

The other type of cells are macrophages with a sample size of 72 cells.
The analysis is done with two groups. A HIV negative (HIV-) control
group and a HIV positive (HIV+) group.

\hypertarget{cell-isolation}{%
\subsubsection{Cell Isolation}\label{cell-isolation}}

\hypertarget{lysis}{%
\subsubsection{Lysis}\label{lysis}}

@@

\hypertarget{labeling-techniques}{%
\subsubsection{Labeling techniques}\label{labeling-techniques}}

For differential analysis proteins need to be labeled to compare mass to
charge intensities in order to quantify observed peptides. Since mass
spectrometry is not a quantitative technique by itself, the peak height
or area does not reflect the abundance of a peptide. Physicochemical
properties of the proteins can change the ionization efficiency and
detectability of the target. However, when comparing the same analyte
between multiple runs of labeled proteins, differences in the mass
spectrum reflect the abundance of those. Labels should be chosen to
change solely the mass of the sample and to not affect folding or other
inherent properties of the protein.

\hypertarget{metabolic-labeling}{%
\paragraph{Metabolic labeling}\label{metabolic-labeling}}

Feeding cells with aminoacids containing heavy isotopes, is the method
of choice in order to label peptides at the earliest possible level.
This atoms can be heavy nitrogen in aminoacids or salts in fertilizer
for plants. Mass shifts are proportional to the isotopes incorporated
during biomass production and are visible after proteolytic cleavage.
Stable isotope labeling in cell culture (SILAC) was presented in the
early 2000s. This method used heavy aminoacid enriched media to feed
cells, in order to quantitatively analyze expression profiles.

\hypertarget{isobaric-labeling}{%
\paragraph{Isobaric labeling}\label{isobaric-labeling}}

\hypertarget{tandem-mass-tag-tmt}{%
\subparagraph{Tandem mass tag (TMT)}\label{tandem-mass-tag-tmt}}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
2cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] { \expandafter\origfigure\expandafter[H] } { \endorigfigure }
\usepackage{sectsty}
\usepackage{paralist}
\usepackage{setspace}\spacing{1.5}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{dcolumn}
\usepackage{natbib}\bibliographystyle{agsm}
\usepackage[nottoc, numbib]{tocbibind}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

With the growing prominence of single-cell techniques across various
omics fields, there is a pressing need to develop a standardized
pipeline for proteomics data in the realm of systems biology. Unlike DNA
sequencing and RNA sequencing, proteomics analysis via mass spectrometry
incurs high costs in terms of labor and equipment. Additionally,
commercially available software solutions often come with hefty price
tags and limited transparency regarding the underlying methods employed.
However, MaxQuant \citep{Cox2008} presents a promising alternative,
complemented by the flexibility using the R programming language.
Introducing ProteoScanR, a state-of-the-art proteomics pipeline
integrated into the user-friendly Proteomics Workbench interface. Guided
by SCoPE2 \citep{Specht2021, Petelski2021, Vanderaa2021}, the
development process of ProteoScanR was thoroughly tested and validated
using both bulk and single-cell mass spectrometry data sets. The
pipeline implementation in the Proteomics Workbench leverages the power
of an interactive environment built in R Shiny, empowering users to
discover valuable insights for their specific data set. Within the
interactive environment, users have the flexibility to customize cutoffs
and thresholds for quality control, as well as employ various approaches
for data transformation, normalization, missing value imputation, and
batch correction. ProteoScanR and the Proteomics Workbench served as a
valuable tool in guiding the identification of significantly expressed
proteins in cells under study. Subsequently, the pathway enrichment
analysis provided additional biological contexts for a comprehensive
understanding of their functional implications. The master thesis
project serves as a foundation for future advancements in the field of
single-cell proteomics. Moreover, the codebase has been designed with
robustness and scalability in mind, ensuring ease of maintenance and
future expansion of the application.
\url{https://github.com/Lukas67/ProteoScanR}

@@

In addition to my faithful feline companion, Apollo, who took care of my
place in Austria while I was working on this project in Stockholm, I am
deeply grateful to all the members of Ujjwal Neogi's research group.

Their unwavering guidance and support throughout this project have been
truly invaluable, and I am profoundly thankful for their expertise and
assistance. \pagenumbering{arabic}

@@

\hypertarget{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}{%
\subsection{From the Central Dogma of Molecular Biology to Omics and
Systems
Biology}\label{from-the-central-dogma-of-molecular-biology-to-omics-and-systems-biology}}

Before the initial exploration of molecular biology, the processes of
life were roughly inferred and not well understood. The central dogma of
molecular biology serves as the cornerstone of biological processes,
providing a framework for understanding how genetic information is
converted into functional proteins. This framework consists of two major
processes: transcription, which converts DNA into RNA, and translation,
which converts RNA into functional proteins. Between these processes,
molecular-level regulation occurs through chemical adaptations. This
regulation is crucial for biological organisms to efficiently produce
their genetic products, known as proteins. Crick's formulation of the
central dogma has guided research in molecular biology for over 60 years
\citep{Cobb2017}. Over the past six decades, researchers have explored
various aspects of this principle, collectively referred to as
``omics,'' to comprehensively analyze biomolecules in diverse contexts.
Omics research has emerged as an approach to study biomolecules and
their interactions within biological systems comprehensively. Omics
encompasses various fields, including genomics, transcriptomics,
proteomics, and metabolomics, with each field focusing on a specific
class of biomolecules. By analyzing multiple layers of omics data,
researchers can gain a holistic view of biological processes, uncover
regulatory mechanisms, and identify key players in complex biological
networks using computational methods. Systems biology takes a
quantitative approach to investigate the interactions, dynamics, and
emergent properties of biomolecular networks. Biological systems are
imagined as complex, multi-layer networks, with the ensemble of
proteins, known as the proteome, playing a critical role in this
interactive structure. Proteins have various functions, such as
providing structural integrity, catalyzing chemical reactions, and
regulating cellular functions \citep{Karahalil2016}

@@

The central dogma of molecular biology offers insights not only at the
level of individual terms but also provides a hierarchical understanding
of biology. Hierarchical organization is the underlying principle in all
topics of biology. Proteins exhibit hierarchical structures that
contribute to their functionality. At the primary structure level,
proteins are composed of smaller building blocks called amino acids.
Combining them results in the protein secondary structure such as alpha
helix and beta sheet. These secondary structures further assemble to
create higher-order tertiary structures, which represent specific
domains within the protein. Finally, the quaternary structure describes
the functional state of the protein as a whole at a given time.
Proteomics is a comprehensive approach that investigates the complete
protein composition of a specimen, aiming to understand the intricate
biological network it represents. For instance if pathogens interact
with cells of an organism a cascade of biochemical reactions takes
place. These reactions influence the outcome of the cellular proteome in
regards of localization, abundance, post-translational modifications
\citep{Beltran2017}. By studying the proteome, we can gain a
comprehensive understanding of the hierarchical organization and
functional dynamics of biological processes.

\hypertarget{bulk-proteomics}{%
\subsubsection{Bulk proteomics}\label{bulk-proteomics}}

Higher organisms are composed of specialized cells organized into
tissues, such as skin, muscle, and blood. Each tissue consists of cells
with specific functions, resulting in variations in protein expression.
Bulk proteomics is a technique used to analyze the protein composition
of a tissue sample, which contains all types of cells present in that
particular tissue. This approach finds valuable applications for
instance in oncology, where it can provide insights into the protein
profile of a tumor. By studying oncogenic biomarkers and expression
patterns, this knowledge can contribute to the development of screening
methods and the customization of treatments. Additionally, bulk
proteomics can aid in identifying the progression of tumor development
\citep{Kwon2021}. Bulk analysis expression profiles provide scientists
with an average protein abundance across all cells in a sample, offering
a comprehensive view of protein expression. Additionally, compared to
single-cell methods, bulk samples typically contain a greater number of
proteins. In fact the expanded library can be used to validate the
performance of single-cell type and single-cell proteomics methods and
bioinformatics pipeline \citep{Schoof2021}. However, in this type of
analysis, the precise properties of individual cell types can only be
inferred and remain obscured.

\hypertarget{single-cell-type-proteomics}{%
\subsubsection{Single-cell type
proteomics}\label{single-cell-type-proteomics}}

Taking tissue samples can lead to an averaging effect across the entire
cellular ensemble, making it difficult to discern specific cell types.
To overcome this limitation, cell sorting techniques such as
fluorescence-activated cell sorting (FACS), magnetic-activated cell
sorting (MACS), and buoyancy-activated cell sorting (BACS) are employed
\citep{Liou2015}. These techniques primarily rely on the detection of
surface proteins on cells and the specific binding of antibodies to
these proteins. Auto-fluorescence is observed in certain cells, wherein
they possess the capability to emit a light signal when stimulated, and
this emission can be used to sort them accordingly. However
auto-fluorescence can lack of accuracy as shown in studies indicating
the effect of laboratory gloves on cellular samples
\citep{Philpott2017}. On the other side antibodies, typically derived
from the immune system of other species, bind to antigens in a highly
specific lock-and-key manner. Antibodies are widely used in molecular
biology and find applications in cell isolation. Depending on the
desired application, the antibodies can be labeled with fluorescent
markers (FACS), magnetic beads (MACS), or biotin (BACS). These methods
have varying technical requirements, with flow cytometers being the most
complex and costly. In MACS, magnetic beads are used to isolate cells
from a solution by utilizing a magnet. However, mechanical forces
involved in MACS can potentially damage the cells. BACS, a method
developed in 2015, employs microbubbles to carry the antibodies for
cellular isolation. The method showed improvements compared to MACS in
regards of viability after sorting. It is important to note that FACS is
considered the gold standard for validating the effectiveness of MACS
and BACS due to its high purity and yield. FACS was used in the
validation of these two methods \citep{Liou2015, Sutermaster2019}.
Single-cell type proteomics reflects the protein ensemble of a specific
cell type at a particular time, providing a focused perspective on the
studied field compared to bulk methods \citep{Maes2020}.

\hypertarget{single-cell-proteomics}{%
\subsubsection{Single-cell proteomics}\label{single-cell-proteomics}}

Cell sorting techniques primarily rely on extracellular proteins to
identify and separate different cell types. However, the intracellular
content of cells can still exhibit variability in experiments.
Single-cell proteomics (SCP) addresses this limitation by providing a
snapshot of the underlying processes within an individual cell,
eliminating this variability and allowing for the elucidation of the
principles of cell type behavior, such as differentiation.
Differentiation, a critical process in various biological systems such
as the immune system, involves the acquisition of specialized functions
by specific cell types \citep{Fang2018}. Investigating differentiation
in multiple cells can be challenging due to the inherent heterogeneity
during the initiation of this process. However, by analyzing one cell at
a time and observing it throughout the differentiation process,
researchers can capture the complete trajectory of differentiation at a
single-cell level. This approach provides valuable insights into the
dynamic changes occurring within cells during differentiation. In 2021,
Specht et al.~developed the SCoPE2 sample preparation and analysis
pipeline, performing macrophage-monocyte differentiation experiments
\citep{Specht2021}. This pipeline streamlined the sample preparation
process, reducing labor time and providing a reference model for
subsequent SCP projects. SCP has proven to be a powerful tool for
studying a wide range of biological phenomena, including drug responses,
infectious diseases, and organism development. Over the past decade,
single-cell techniques, including SCP, have become indispensable tools
for the global research community \citep{Minakshi2019}.

\hypertarget{capturing-the-proteome}{%
\subsection{Capturing the proteome}\label{capturing-the-proteome}}

An early approach of qualitative analysis of the cellular proteome
involved labeling with fluorescent antibodies and imaging. The major
disadvantage of this technique was the limitation to only stain a few
proteins per cell and the requirement of a-priori knowledge of the
protein to analyze. For quantification, procedures such as Western
blots, immunoassays or cytometry by time of flight (CyTOF) have been
used. Challenges are the ability to permeate cells, accessibility and
binding of the epitope and the creation of specific antibodies for a
given protein \citep{Budnik2018}. Early quantitative methods in
molecular biology, measuring the entire protein concentration, were
assays like the UV-vis-, Bradford- and/or bicinchoninic-acid-assay.
These methods are still in use to obtain total concentration for sample
preparation. Another quantitative technique involved RNA-sequencing.
RNAseq refers to the task of examining the transcriptome of bulk tissue
sample with deep-sequencing technologies. RNA in a sample includes
mRNAs, non-coding RNAs, and small RNAs. One of the main disadvantages
is, that the mapping relies on the knowledge of the DNA sequence of the
observed organism (reference genome) \citep{Wang2009}. Another
disadvantage is, that the amount of protein translated from the
determined mRNA can only be estimated \citep{Gygi1999}. However, today
these predictions can be done with computational methods such as LASSO
and the accuracy of the estimate highly depends on selection and
validation of this method \citep{Magnusson2022}. When observing the
protein ensemble of a cellular environment with a top-down approach like
DNAseq or RNAseq, the characteristics (e.g.~post-translational
modifications or charge) and local concentration of the proteins remain
hidden. The cause of this phenomenon is that only a fraction of the DNA
is transcribed into mRNA and not all mRNA is further translated into
protein. Furthermore multiple proteins can be derived from a single
DNA-sequence with alternative splicing and different proteins coming
from non related sequences acting like building blocks can form protein
complexes and/or Protein-RNA complexes. Cells are highly efficient and
do not waste energy by producing obsolete proteins, moreover the
degradation of proteins in the proteasome is an important energy
recovery mechanism \citep{TANAKA2009}. A bottom-up strategy can
elucidate these properties of the protein ensemble at a given time
leading to proteomics. When studying impact of an external signal or
stressor to a cell the abundance and properties of a protein or
protein-ensemble can enlighten the underlying pathway leading to a
biological response. Since some proteins do not even function before
post-translational modifications (PTMs) happen, a knowledge about these
can be crucial for further studies. Mass spectrometry (MS) fusions the
advantages of the above mentioned techniques, because it combines the
specificity of qualitative assays which are even able to sense (PTMs)
with the quantitative accuracy of protein concentration assays.
Furthermore MS does not require any a-priori knowledge of the protein
ensemble of the specimen, making it the method of choice for untargeted
approaches.

\hypertarget{mass-spectrometry}{%
\subsubsection{Mass Spectrometry}\label{mass-spectrometry}}

Mass spectrometry enables qualitative and quantitative analysis of the
entire repertoire of a biological sample. Mass spectrometers measure the
mass to charge ratio (m/z) of a particle from a fragmented larger
molecule. This is achieved by a physical procedure done with a device
which is made of three major components: the ion source, an analyzer,
and a detector. The ion sources charges molecules and accelerate them
through a magnetic field. The analyzer separates particles according to
their mass to charge ratio and the detector senses charged particles and
amplifies their signal \citep{Parker2010}. However before acquiring any
data, sample preparation and pre-processing is required. After sample
preparation the first step involves tryptic digestion. It is a widely
used technique in shotgun proteomics and involves the enzymatic cleavage
of proteins into smaller peptides using the protease enzyme trypsin. The
process known as proteolysis takes place in multiple parts of multi
cellular organisms and was firstly detected in the small intestine
digesting proteins \citep{Wang2008}. The enzyme cleaves specifically at
the caroboxyl side of arginine and lysin residues. Afterwards a charge
remains on the residues making them detectable for the mass
spectrometer. It is proven that proteins which are digested to the
limit, meaning no further proteolysis can be induced, provide the best
results in such bottom up approaches \citep{Brownridge2011, Laskay2013}.

To observe multiple samples in one run, labeling is needed to identify
each sample within a batch. Since mass spectrometry is not a
quantitative technique by itself, the peak height or area does not
reflect the abundance of a peptide. Physicochemical properties of the
peptides can change the ionization efficiency and detectability of the
target. However, when comparing the same analyte between multiple runs
of labeled peptides, differences in the reporter ion intensity reflect
the abundance of those. Labels should be chosen to change solely the
mass of the sample and to not affect folding or other inherent
properties of the peptide. The two techniques for labeling peptides are
metabolic labeling and isobaric labling. In metabolic labeling the cells
were fed with aminoacids containing heavy isotopes, it is the method of
choice in order to label peptides at the earliest possible level. This
atoms can be heavy nitrogen in aminoacids or salts in fertilizer for
plants \citep{Karen2012}. Mass shifts are proportional to the isotopes
incorporated during biomass production and are visible after proteolytic
cleavage. Stable isotope labeling in cell culture (SILAC) was presented
in the early 2000s. This method used heavy aminoacid enriched media to
feed cells, in order to quantitatively analyze expression profiles. The
limitation of this method is, that it can only be applied to a human
cell culture (in vitro) or to model organisms (in vivo). Isobaric
labeling solely applies to in vitro techniques, with TMT and iTRAQ. The
data analyzed by ProteoScanR primarily utilized TMT tags. Tandem mass
tag (TMT) reagents enable differentiation of multiple samples in a
single mass spectrometry (MS) run. Each sample is individually labeled
with TMT reagents and then pooled together, this procedure is called
multiplexing. TMTs have the same charge and differ only by their
isotopic masses, the peaks observed for each sample are called reporter
ions (RI). Each RI and sample is interpreted as one channel in
downstream analysis. This technique allows for the identification,
enrichment, and quantification of low-abundance peptide ions, which is
particularly valuable in single-cell techniques. With this technique it
is possible to quantify proteins and differ low abundant proteins from
background noise. However, a drawback of isobaric labeling is the
presence of co-fragmentation signals in the spectrogram, which requires
data normalization to remove unwanted contributions
\citep{Marx2019, Budnik2018}. Isotopic distribution in TMTs, reflecting
the natural distribution, can be corrected during data acquisition by
considering a defined spread in other channels. Single-cell proteomics
by mass spectrometry (SCoPE-MS) uses a isobaric labeling with a tandem
mass tag (TMT) to enhance signal intensity of a protein species. Loading
of a single-cell channel (SCC) with the right amount can be balanced to
the protein concentration of a single cell \citep{Ye2022}.

Before ionization peptides need to be further separated according to
their chemical properties, size or species by liquid chromatography
(LC), because the number of peptide processed simultaneously by the MS
device is limited. In the case of mass spectrometry, liquid
chromatography especially high performance LC (HPLC) is the technique of
choice and replaces the instrumentation-wise simpler technique gel
electrophoresis. In HPLC the molecules eluate through a narrow column
with high pressures (50-350 bar), compared to LC where the separation is
gravity dependent. Similar to gel electrophoresis molecules can be
separated upon different chemical properties, depending on the type of
the chromatography column. In a typical proteomics experiment peptides
are separated with reversed phase chromatography (RPLC). The principle
behind RPLC is a column with a nonpolar stationary phase which interacts
with the nonpolar residues of the peptides. Peptides were fractionated
according to their hydrophobicity \citep{Pitt2009}.

To analyze a biological sample consisting of peptides in solution the
liquid needs to be vaporized into gas phase. Two techniques are capable
of this procedure. Electrospray ionization (ESI) pushes the analyte
through a capillary while applying an electric current to the liquid,
vaporizing the sample to a charged aerosol. Peptides are further
fragmented according to their chemical properties and can be further
handled in the mass spectrometer. The fragmented peptides are now in
charged droplets separated based on their surface charge, splitting
further into smaller droplets until they reach the gas phase as ions.
Two physical models describe the process from gas phase to ion called
``The ion evaporation model'' (IEM) and ``The charge residue model''
(CRM). In the ion evaporation model the droplets shrink by evaporation
until ions are expelled \citep{Iribarne1976}. The model had its
limitation by explaining same evaporation rate constant among ions with
different chemical properties. In the charge residue model the
assumption of one molecule per droplet leads to an ionization rate
constant, which is independent of the ion itself and relies solely on
the generation of the droplet and the efficiency of the solvent
\citep{Wilm2011}.

Subsequently after the ionization peptides are accelerated through the
magnetic field generated by the mass analyzer. Six general types of mass
analyzers are availaible on the market, there are Quadropole, time of
flight (TOF), magnetic sector, electrostatic sector, quadropole ion trap
and ion cyclotron resonance -mass analyzer. The function of the mass
analyzer is to separate ions based on their mass to charge ratio (m/z).

Before the signal of the separated ions can be obtained an optional
subsequent step involves trapping the ions an electric and/or magnetic
field in order to detect ions based on their (m/z) consecutively. Ion
trapping enables the technique of coupled mass spectrometers, where ions
were further fragmented by collision induced dissociation and analyzed
subsequently. The detector transforms the impact energy of the ion into
an electric signal, called spectrum which can be processed
computationally.

Given the high resolution of MS data, algorithms are employed to convert
the raw signal into an interpretable form. Software packages like
MaxQuant \citep{Cox2008} are commonly used to process the data,
providing it for further analysis and statistical testing. Other
software solutions include Proteome Discoverer by Thermo Fisher.

The data obtained from mass spectrometry has three dimensions: m/z
ratio, intensity, and retention time. To separate peaks from each other,
algorithms are used to identify local minima in the data. The centroid
of each peak is determined by fitting a Gaussian peak shape, which can
be interpreted as locating the peaks of each m/z spectrum as a function
of retention time. The centroid of a peak corresponds to an isotope.

In order to decipher the isotopic distribution of a biomolecule,
MaxQuant employs a vertex-based approach. This process, referred to as
de-isotoping, utilizes graph theory. It creates a vertex for each
individual peak and connects them with their potential isotopic
counterparts. This is achieved by determining the mass proportion of an
average amino acid to its respective isotope using a concept known as
averagine \citep{Senko1995}. By applying this procedure, the number of
data points is reduced by a factor of ten, and each peak represents a
small biomolecule.

The subsequent step in data acquisition involves the detection of labels
for quantification purposes. Isotopic pairs of the label (such as N13,
N14, N15) present in the tag or aminoacid are identified by convolving
the two measured isotope patterns with the theoretical isotope patterns.
Through an iterative process using a least-squares method, the best fit
is determined, enabling the identification of the specific channel or
sample.

The intensity-weighted average of the MS peak centroids, as described as
3D peak identification, corresponds to the mass of the peptide. The
accuracy of mass measurements relies on the specific analyzer used. For
an Orbitrap-type analyzer (which was used during this project), MaxQuant
applies a correction value of 1 ppm (parts per million). Autocorrelation
between centroids is accounted for by considering only well-identified
peptides. Published data indicates that the mass precision within an MS
experiment typically ranges around 10\^{}-7 \citep{Wilschefski2019}.

Each peptide is represented by its individual fingerprint in the MS
spectrum, which is based on the sequence and modifications of amino
acids. The m/z ratio of the amino acids can be calculated, and the
resulting peaks are interpreted as an amino acid sequence. However, only
peptide fragments are visible in the spectrum due to peptide
fragmentation. To identify proteins, peptides are matched against a
sequence database. Sequence databases are typically stored as .fasta
files and can be downloaded from the UniProt database (www.uniprot.org).
The availability of gene sequences in databases and the computational
matching of peptides against those sequences enable the identification
of alterations in a sample at the protein level. These alterations can
rely on the sequence level or could be to post-translational
modifications (PTMs) such as phosphorylations or methylations
\citep{Aebersold2003}. The peptide identification (P-) score reflects
the goodness of fit between the data and the identified sequence in the
database, taking into account the length of the peptide. This score is
used to calculate the posterior error probability, which is then
utilized to estimate the false discovery rate (FDR). The FDR calculation
provides a measure of the rate at which false identifications are
expected in the dataset. Within the identified peptide groups, a ``razor
protein'' refers to the protein with the highest number of ambiguously
identified peptides. Quantification is performed by considering only
unique peptides for comparison. Posterior error probabilities, which
represent the likelihood of the identified peptide being a random event,
are multiplied together, and only distinct sequences with the highest
scores are considered for further analysis. Various metrics that
indicate the performance of the peptide search can be utilized in
downstream analysis.

\hypertarget{downstream-analysis}{%
\subsubsection{Downstream analysis}\label{downstream-analysis}}

With advancements in mass spectrometry technology and computational
methods, proteomic analysis has emerged as a powerful tool for
investigating complex protein samples. However, analyzing proteomic data
poses challenges in data processing, statistical analysis, and
interpretation. This thesis aims to address these challenges by
exploring computational methods for downstream analysis of proteomics
data. The methods employed in this study will be thoroughly explained in
the methods section, with a specific focus on developing an interactive
environment equipped with a graphical user interface, called Proteomics
Workbench. This user-friendly interface will enable researchers,
including those with limited computational expertise, to navigate and
comprehend the data effectively.

ProteoScanR comprises a series of steps to streamline the data. It
begins with quality control measures at the peptide spectrum match
level, involving the removal of contaminants and irrelevant channels.
Significance levels of peptide spectrum matches serve as confidence
thresholds for match selection. The data is then aggregated at the
peptide level, followed by additional filtering procedures based on
metrics such as reporter ion intensity, peptide coefficient of variation
over razor proteins, and missing rates for peptides.

To derive meaningful biological insights from the dataset, statistical
testing plays a crucial role. Selecting appropriate statistical tests
and ensuring proper data transformation and normalization are essential
steps. Many statistical methods rely on data distribution and require
careful consideration to preserve the biological interpretation of the
results. This thesis employs various transformation and normalization
techniques, aiming to maintain data integrity and avoid biased
interpretations. Additionally, an entropy-based visualization approach
will provide users with a valuable tool for validating the computations
and selected thresholds for the filtering steps.

Handling missing values poses further challenges in the pipeline
development since a missing signal for a peptide results in a missing
protein, which is highly unlikely in a biological context. Several
options for missing value handling will be included and thoroughly
explained in the methods section.

Dimensionality reduction and visualization techniques are widely used to
cluster data. In this project, multiple approaches will be utilized, and
users will have the ability to visually and interactively observe the
data.

It is important to acknowledge that downstream analysis in proteomics
lacks a standardized approach, and the selection of computational
methods depends on the specific dataset and research objectives.
Furthermore, statistical analysis and the interpretation of biological
meaning often present misconceptions that need to be addressed. For
instance, batch correction, a commonly practiced method, has been
overestimated in numerous studies, leading to misleading results
\citep{Nygaard2016}. This thesis aims to contribute to the field by
providing researchers with a comprehensive understanding of
computational methods in proteomics analysis and their application in an
interactive environment.

@@

\hypertarget{wetlab---materials-and-methods}{%
\subsection{Wetlab - Materials and
Methods}\label{wetlab---materials-and-methods}}

For testing and development of the ProteoScanR pipeline the following
materials and methods were performed. For the sample preparation
protocol see appendix.

\hypertarget{celltypes}{%
\subsubsection{Celltypes}\label{celltypes}}

Macrophages from 12 healthy donors and 12 HIV patients with metabolic
syndrome and 12 HIV patients without metabolic syndrome. Cell sorting
was performed with FACSAria Fusion (BD Biosciences).

@@

Protein digestion was performed using trypsin (Promega, Madison, WA)
buffered in Triethylammonium bicarbonate buffer (TEAB). Reagents were
handled with MANTIS automatic dispenser (Formulatrix, Bedford, MA).

\hypertarget{labeling}{%
\subsubsection{Labeling}\label{labeling}}

Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher)
in anhydrous acetonitril (ACN) solution.

\hypertarget{pooling}{%
\subsubsection{Pooling}\label{pooling}}

Carrier proteome was added to the 36 TMTpro sets before pooling sample
types for each of the 36 MS-runs.

\hypertarget{liquid-chromatography}{%
\subsubsection{Liquid chromatography}\label{liquid-chromatography}}

In order to separate peptides according to their chemical properties,
size or species a liquid chromatography (LC) is recommended before
ionization. For the reference data used for developing the pipeline the
following intrumentation was used. EASY-Spray™ HPLC column, ES802A, 2
\(\mu\)m × 75 \(\mu\)m ID × 250 mm (ThermoFisher Scientific) Acclaim
PepMap trap column, 2 \(\mu\)m × 75 \(\mu\)m ID × 2 cm (ThermoFisher
Scientific)

\hypertarget{mass-spectrometry-1}{%
\subsubsection{Mass Spectrometry}\label{mass-spectrometry-1}}

\hypertarget{ionization}{%
\paragraph{Ionization}\label{ionization}}

For the reference data ESI ionization was used.

\hypertarget{mass-spectrometer}{%
\paragraph{Mass spectrometer}\label{mass-spectrometer}}
